---
author: "Abigail Lebrecht"
date: 2018-05-07
title: "The local elections in Barnet: What really happened?"
tags: ["election","open data","R"]
---
The Barnet council result has been the most startling in the recent local elections. Consistently predicted in  pre-election polls as a Labour gain, Labour's top target ended up with an an increased majority for the Conservatives. While the London pre-election polls were widely accurate, Barnet was the only area which bucked the trends. Many local people who were on the campaign trail have been quick to connect the Barnet result with on Labour's anti-semitism problems including [Labour councilors](https://news.sky.com/story/defeated-labour-candidate-people-think-were-racist-in-wake-of-anti-semitism-row-11361324) and [Barnet's Labour leader](https://www.theguardian.com/commentisfree/2018/may/05/barnet-paid-price-for-labour-record-on-antisemitism). A [blog by Dr Daniel Allington](http://www.danielallington.net/2018/05/the-jewish-vote-in-barnet/) provides evidence for the link by showing a positive correlation between the increase in Conservative vote share in 2018 and the proportion of a ward that is Jewish and a similar decrease in the Labour vote, inferring that anti-semitism in Labour is likely to have driven the result in these key wards.

However, there are those who downplay this conclusion. As in the 2017 General Election, there was a swing to both Labour and the Conservatives with Labour increasing their vote, both in total amount of votes and swing, so how is that a rejection of Labour? Some who dispute the impact of anti-semitism use demographic studies to point out that many Jewish voters already vote Conservative, implying that a decrease in Jewish people voting Labour would not have a huge impact. Finally, some mention the collapse of UKIP since 2014 as a factor that may have helped the Conservatives in Barnet, as it has done nationally. The UKIP argument is an odd one. In 2014, UKIP polled less than 2% of total votes in Barnet. Such was their failure that they only bothered fielding one candidate in the whole of Barnet in 2018, thus a 'swing' of 2% away from them. This tiny number of UKIP voters spread across all wards were too disparate to have any clear impact on the Barnet election.

The detractors are correct that Labour did increase their votes and swing in Barnet in 2018, but not as much as the Conservatives. So what really happened in Barnet and do the vote numbers support the anecdotal evidence from campaigners that anti-semitism mattered in key areas?

It is common to analyse election results by vote shares and swings. This gives the impression that the Conservatives could only have won votes from those who previously voted for other parties (leading to the bizarre UKIP theory). In reality, turnout at local elections is historically low and voters come and go. It is easier to convince a Conservative supporter who isn't planning on voting to go out and vote, then a Labour supporter to vote Conservative. The [British Election Study](http://www.britishelectionstudy.com/bes-resources/2017-election-bes-internet-panel-data-released) for 2017, showed that Labour won many more new voters from people who hadn't voted in 2015, than from any single party. Their success was formed in bringing out the vote more than changing people's minds (the Brexit referendum helped with this engagement). Thus, to understand what happened in Barnet, we need to understand those who don't always vote as well as those who do. Throughout, we consider voting figures as the combined vote for each party (all voters had up to three votes). Hence votes are not equivalent to voters. First we plot total votes for each ward in Barnet in 2014 and 2018:

```{r warning=FALSE,echo=FALSE,message=FALSE}
library(httr)
library(XML)
library(tidyverse)
library(magrittr)

url <- "https://en.wikipedia.org/wiki/Barnet_London_Borough_Council_election,_2018"

r <- GET(url)

doc18 <- readHTMLTable(
  doc=content(r, "text"),header = TRUE,stringsAsFactors = FALSE)

url <- "https://en.wikipedia.org/wiki/Barnet_London_Borough_Council_election,_2014"

r <- GET(url)

doc14 <- readHTMLTable(
  doc=content(r, "text"),header = TRUE,stringsAsFactors = FALSE)

totalresults = full_join(doc14$`Barnet Local Election Result 2014[5]`[,c(2,9)],doc18$`Barnet Local Election Result 2018`[,c(2,9)],by = 'V2')
totalresults[is.na(totalresults)]=0  
totalresults[,2:3] %<>% lapply(function(x) as.numeric(gsub(",", "", x,fixed=TRUE)))
names(totalresults) = c('Party','2014','2018')

#totalresults %>% gather(year,votes,-Party) %>% ggplot()+geom_point(aes(x=year,y=votes,colour=Party))+theme_bw()

wards14 =  do.call("rbind.data.frame",(lapply(doc14[grepl('(3 seats)',names(doc14))] ,function(x){data.frame(x[['V2']],x[['V4']])}))) 

wards14= data.frame(ward=sapply(row.names(wards14),function(x){unlist(strsplit(x,'(',fixed = TRUE))[1]}), party=wards14[,1],votes=wards14[,2],year=rep(2014,nrow(wards14))) %>% filter(votes!="")

wards18 =  do.call("rbind.data.frame",(lapply(doc18[names(doc18)!='NULL'][-(1:3)] ,function(x){data.frame(x[['V2']],x[['V4']])}))) 

wards18 = data.frame(ward=sapply(row.names(wards18),function(x){unlist(strsplit(x,'.',fixed = TRUE))[1]}), party=wards18[,1],votes=wards18[,2],year=rep(2018,nrow(wards18))) %>% filter(votes!="")

wards = rbind(wards14,wards18) %>% mutate_all(as.character)
wards$ward = trimws(wards$ward)
wards$votes =  as.numeric(gsub(",", "", wards$votes,fixed=TRUE))
wards[!(wards$party %in% c('Conservative','Labour','Liberal Democrat','Green','UKIP')),]$party='Other'
wards = wards %>% group_by(ward,party,year) %>% summarise(votes = sum(votes)) %>% group_by(ward,year) %>% mutate(total=sum(votes))


ggplot(wards,aes(x=year,y=votes,colour=party))+geom_point()+theme_bw()+geom_line(aes(group=party))+facet_wrap(~ward,nrow=3)+ scale_colour_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+theme(strip.text.x = element_text(size = 6,hjust=0),legend.text = element_text(size=7))
                                                                                                  
```

In real vote numbers we see the Conservatives' vote went up in every ward, dramatically so in many safe wards as well as some marginal (e.g. Hale). Labour's vote was more sporadic, with their most spectacular increase in High Barnet and more modest increases or decreases in other wards. While the UKIP vote in 2014 was negligible, the Green party did surprisingly well in 2014, but fielded less candidates in 2018. As in the 2017 General Election, we see Labour as the main beneficiary from the 2014 Green vote with Labour rises in Coppets, High Barnet, Woodhouse, East Finchley, East Barnet and Underhill correlating with the more significant 2014 Green votes there.  

Some of the vote increases for the Conservatives (and Labour in High Barnet and Colindale) suggest many more votes cast in 2018 than 2014. So how much were Conservative gains in 2018 from previous Labour voters, or did they benefit primarily from turnout changes? To understand this better we compare the change in total number of votes for each party between 2014 and 2018 with the overall change of total votes cast in each ward between the two elections.
                                                                                                  

```{r warning=FALSE,echo=FALSE,message=FALSE}
sprwards = wards %>% select(-total) %>%  spread(year,votes,convert=TRUE,fill=0) %>%  mutate(increase = (`2018`-`2014`))

twards = wards %>% group_by(ward,year) %>% summarise(votes = sum(votes)) %>% spread(year,votes,convert=TRUE) %>% mutate(turnoutchange = (`2018`-`2014`))

changes = inner_join(select(sprwards,ward,party,increase),select(twards,turnoutchange))

ggplot(changes,aes(x=increase,y=turnoutchange,colour=party))+geom_point()+geom_text(aes(label=ward),size=3,hjust=0,nudge_x = 20)+theme_bw()+ scale_colour_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+ggtitle("Change in party votes against change in total votes for each ward")+xlab("Change in party vote, 2018 to 2014")+ylab("Change in total votes, 2018 to 2014")
```

While, this graph looks a bit of a mess of coloured points, one trend immediately catches the eye - the Conservative points suggest a linear trend: a relationship between increase in turnout and increase in the Conservative vote. We can investigate this further by replotting this and putting in linear regression lines and error bars:
                                                                                                  

```{r warning=FALSE,echo=FALSE,message=FALSE}
ggplot(changes,aes(x=increase,y=turnoutchange,colour=party))+geom_point()+geom_smooth(aes(group=party,fill=party),method="lm",alpha=0.1)+theme_bw()+ scale_colour_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+ scale_fill_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+ggtitle("Change in party votes against change in total votes for each ward")+xlab("Change in party vote, 2018 to 2014")+ylab("Change in total votes, 2018 to 2014")
```

Our linear regression lines show that only the Conservative vote had a clear, steep, relationship with increase in turnout, with the narrowest error bars of all parties. It suggests that the Conservative's success was down to mobilising dormant voters in key wards (many of which have high Jewish populations), possibly more than (but in addition to) changing minds of other party's voters. This doesn't mean that Labour didn't benefit from increased turnout. In High Barnet, Oakleigh, Childs Hill and Colindale, Labour appear to have attracted new voters since 2014. Yet this success is sporadic rather than a trend across Barnet. 

This undermines the demographic argument that the anti-semitism issue could not have made a discernable impact. Not only were Labour voters voting for other parties or staying at home because of anti-semitism, as reported by Labour canvassers, but possibly an even greater number of dormant voters came out to vote Conservative probably for this reason too. This is supported by reports from a [Lib Dem candidate in Mill Hill](http://barneteye.blogspot.co.uk/2018/05/barnet-council-elections-2018-in-depth.html) suggesting that increased turnout was crucial, to quote: '*In short, the Barnet Conservatives got every last one of their staunch voters out, which is completely unprecendented in Barnet local elections. *' The combination of vote switching and increased turnout handed the Conservatives thousands of votes across Barnet and made the difference.

What do these findings tell us about why Labour did so badly in Barnet in 2018? Voter turnout increases when voters are carefully targetted and have something they are passionate about to vote for, or vote against. In 2016 and 2017, we saw record turnout nationally after years of low turnouts, as Brexit divided people and simultaneously politically engaged them. Barnet's increase in turnout and surge to the Conservatives in 2018 contradicted UK trends for regional and demographic voting in remain supporting areas. In 2017 all three Barnet constituencies saw increased turnout with increased Labour votes, since turnout was up in almost every ward in 2018, surely some of these new engaged voters from 2016 or 2017 were voting again in 2018, thus making the Conservative gains even more remarkable.

It is impossible to say if one key issue turned an election, or to trivialise a result into a set of numbers or a single graph. The Conservatives success can not be described only as attracting votes from Labour or UKIP, nor Labour's failure excused by their increase in votes. But combining the voting patterns and turnout changes with demographic analysis and anecdotal evidence from those on the campaign trail creates a picture that suggests anti-semitism may have been the over-riding issue that drove voters to the Conservatives in many key Barnet wards in 2018.
