---
author: "Abigail Lebrecht"
date: 2018-05-07
title: "The local elections in Barnet: What really happened?"
tags: ["election","open data","R"]
---
The Barnet council result has been the most startling in the recent local elections. Consistently predicted in  [pre-election polls as a Labour gain](https://www.theguardian.com/politics/2018/apr/26/londons-flagship-tory-boroughs-may-be-out-of-labour-reach-poll-predicts), Labour's top target delivered an increased majority for the Conservatives. Many local people from all parties who were on the campaign trail have been quick to connect the Barnet result with Labour's anti-semitism problems including [Labour councillors](https://news.sky.com/story/defeated-labour-candidate-people-think-were-racist-in-wake-of-anti-semitism-row-11361324) and [Barnet's Labour leader](https://www.theguardian.com/commentisfree/2018/may/05/barnet-paid-price-for-labour-record-on-antisemitism). A [blog by Dr Daniel Allington](http://www.danielallington.net/2018/05/the-jewish-vote-in-barnet/) provides evidence for the link by showing a positive correlation between the increase in Conservative vote share in 2018 and the proportion of a ward that is Jewish and a similar decrease in the Labour vote, inferring that anti-semitism in Labour is likely to have driven the result in these key wards.

However, those who defend Labour's performance in Barnet downplay this relationship. They note that Labour increased their vote in Barnet in 2018, both in total amount of votes and swing, thus Labour were endorsed not rejected in Barnet. The main weakness in this argument is that the increase in the Labour vote is representative of national trends since 2015 away from smaller parties (particularly UKIP) and towards pragmatic votes for the two main parties. We saw these trends in the 2017 General Election and they have continued in every poll and most of the local election results. By itself this does not prove or disprove anything about the Barnet vote. They also suggest that the Conservatives' increase can be explained by the collapse of UKIP. UKIP was a tiny force in Barnet in 2014 with less than 2% of the vote. Its drop only appears large because they chose to field only one candidate in 2018 giving the impression of a total collapse. It is assumed that all UKIP voters immediately switch to the Conservatives, but that is a generalisation. The UKIP vote is likely to have split between a few parties and not voting at all and given the small numbers, unlikely to have had any significant impact in most wards in Barnet. Finally some people suggest that many Jewish voters already vote Conservative, implying that a decrease in Jewish people voting Labour would not have a huge impact. 

It is impossible to say if one key issue turned an election, or to trivialise a result into a set of numbers or a single graph. However gathering as much evidence can produce a clearer picture of voting patterns and understand how much credence can be given to the above arguments. In the following blog I will try and delve into the Barnet results and improve the picture on how people voted there.

It is common to analyse election results by vote shares and swings. This gives the impression that the Conservatives could only have won votes from those who previously voted for other parties. In reality, turnout at local elections is historically low and voters come and go. It is easier to convince a Conservative supporter who isn't planning on voting to go out and vote, then a Labour supporter to vote Conservative. The [British Election Study](http://www.britishelectionstudy.com/bes-resources/2017-election-bes-internet-panel-data-released) for 2017, showed that Labour won many more new voters from people who hadn't voted in 2015, than from any single party. Their success was formed in bringing out the vote more than changing people's minds (the Brexit referendum helped with this engagement). Thus, to understand what happened in Barnet, we need to understand those who don't always vote as well as those who do. Throughout, we consider voting figures as the combined vote for each party (all voters had up to three votes). Hence votes are not equivalent to voters. First we plot total votes for each ward in Barnet in 2014 and 2018:

```{r warning=FALSE,echo=FALSE,message=FALSE}
library(httr)
library(XML)
library(tidyverse)
library(magrittr)

url <- "https://en.wikipedia.org/wiki/Barnet_London_Borough_Council_election,_2018"

r <- GET(url)

doc18 <- readHTMLTable(
  doc=content(r, "text"),header = TRUE,stringsAsFactors = FALSE)

url <- "https://en.wikipedia.org/wiki/Barnet_London_Borough_Council_election,_2014"

r <- GET(url)

doc14 <- readHTMLTable(
  doc=content(r, "text"),header = TRUE,stringsAsFactors = FALSE)

totalresults = full_join(doc14$`Barnet Local Election Result 2014[5]`[,c(2,9)],doc18$`Barnet Local Election Result 2018`[,c(2,9)],by = 'V2')
totalresults[is.na(totalresults)]=0  
totalresults[,2:3] %<>% lapply(function(x) as.numeric(gsub(",", "", x,fixed=TRUE)))
names(totalresults) = c('Party','2014','2018')

#totalresults %>% gather(year,votes,-Party) %>% ggplot()+geom_point(aes(x=year,y=votes,colour=Party))+theme_bw()

wards14 =  do.call("rbind.data.frame",(lapply(doc14[grepl('(3 seats)',names(doc14))] ,function(x){data.frame(x[['V2']],x[['V4']])}))) 

wards14= data.frame(ward=sapply(row.names(wards14),function(x){unlist(strsplit(x,'(',fixed = TRUE))[1]}), party=wards14[,1],votes=wards14[,2],year=rep(2014,nrow(wards14))) %>% filter(votes!="")

wards18 =  do.call("rbind.data.frame",(lapply(doc18[names(doc18)!='NULL'][-(1:3)] ,function(x){data.frame(x[['V2']],x[['V4']])}))) 

wards18 = data.frame(ward=sapply(row.names(wards18),function(x){unlist(strsplit(x,'.',fixed = TRUE))[1]}), party=wards18[,1],votes=wards18[,2],year=rep(2018,nrow(wards18))) %>% filter(votes!="")

wards = rbind(wards14,wards18) %>% mutate_all(as.character)
wards$ward = trimws(wards$ward)
wards$votes =  as.numeric(gsub(",", "", wards$votes,fixed=TRUE))
wards[!(wards$party %in% c('Conservative','Labour','Liberal Democrat','Green','UKIP')),]$party='Other'
wards = wards %>% group_by(ward,party,year) %>% summarise(votes = sum(votes)) %>% group_by(ward,year) %>% mutate(total=sum(votes))


ggplot(wards,aes(x=year,y=votes,colour=party))+geom_point()+theme_bw()+geom_line(aes(group=party))+facet_wrap(~ward,nrow=3)+ scale_colour_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+theme(strip.text.x = element_text(size = 6,hjust=0),legend.text = element_text(size=7))
                                                                                                  
```

In real vote numbers we see the Conservatives' vote went up in every ward, dramatically so in many safe wards as well as some marginal (e.g. Hale). Labour's vote increases were more sporadic, with their most spectacular increase in High Barnet and more modest increases or decreases in other wards. The Green party did surprisingly well in 2014, but fielded less candidates in 2018. As in the 2017 General Election, we see Labour as the main beneficiary from the 2014 Green vote with Labour rises in Coppets, High Barnet, Woodhouse, East Finchley, East Barnet and Underhill correlating with the more significant 2014 Green votes there.  

Some of the vote increases for the Conservatives (and Labour in High Barnet and Colindale) suggest much higher turnout in 2018 than 2014. So how much were Conservative gains in 2018 from previous Labour voters, or did they benefit primarily from turnout changes? To understand this better we compare the change in total number of votes for each party between 2014 and 2018 with the overall change of total votes cast in each ward between the two elections.
                                                                                                  

```{r warning=FALSE,echo=FALSE,message=FALSE}
sprwards = wards %>% select(-total) %>%  spread(year,votes,convert=TRUE,fill=0) %>%  mutate(increase = (`2018`-`2014`))

twards = wards %>% group_by(ward,year) %>% summarise(votes = sum(votes)) %>% spread(year,votes,convert=TRUE) %>% mutate(turnoutchange = (`2018`-`2014`))

changes = inner_join(select(sprwards,ward,party,increase),select(twards,turnoutchange))

ggplot(changes,aes(x=increase,y=turnoutchange,colour=party))+geom_point()+geom_text(aes(label=ward),size=3,hjust=0,nudge_x = 20)+theme_bw()+ scale_colour_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+ggtitle("Change in party votes against change in total votes for each ward")+xlab("Change in party vote, 2018 to 2014")+ylab("Change in total votes, 2018 to 2014")
```

While, this graph looks a bit of a mess of coloured points, one trend immediately catches the eye - the Conservative points suggest a linear trend: a relationship between increase in turnout and increase in the Conservative vote. We can investigate this further by replotting this and putting in linear regression lines and error bars:
                                                                                                  

```{r warning=FALSE,echo=FALSE,message=FALSE}
ggplot(changes,aes(x=increase,y=turnoutchange,colour=party))+geom_point()+geom_smooth(aes(group=party,fill=party),method="lm",alpha=0.1)+theme_bw()+ scale_colour_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+ scale_fill_manual(values=c('Conservative'='blue','Green'='green','Labour'='red','Liberal Democrat'='orange','UKIP'='purple','Other'='grey'),name="Party")+ggtitle("Change in party votes against change in total votes for each ward")+xlab("Change in party vote, 2018 to 2014")+ylab("Change in total votes, 2018 to 2014")
```

Our linear regression lines show that only the Conservative vote had a clear, steep, relationship with increase in turnout and had the narrowest error bars of all parties. It suggests that the Conservative's success was down to mobilising dormant voters in key wards (many of which have high Jewish populations), possibly more than (but in addition to) changing minds of other party's voters. This doesn't mean that Labour didn't benefit from increased turnout. In High Barnet, Oakleigh, Childs Hill and Colindale, Labour appear to have attracted new voters since 2014. Yet this success is localised rather than a trend across Barnet. 

This undermines the demographic argument that the anti-semitism issue could not have made a discernable impact. Not only were Labour voters voting for other parties or staying at home reportedly because of anti-semitism, but possibly an even greater number of dormant voters came out to vote Conservative probably for this reason too. This is supported by reports from a [Lib Dem candidate in Mill Hill](http://barneteye.blogspot.co.uk/2018/05/barnet-council-elections-2018-in-depth.html) suggesting that increased turnout was crucial, to quote: '*In short, the Barnet Conservatives got every last one of their staunch voters out, which is completely unprecendented in Barnet local elections. *' The combination of vote switching and increased turnout handed the Conservatives thousands of votes across Barnet and made the difference.

What do these findings tell us about why Labour failed in Barnet in 2018? Voter turnout increases when voters are carefully targetted and have something they are passionate about to vote for, or vote against. In 2016 and 2017, we saw record turnout nationally after years of low turnouts, as Brexit divided people and simultaneously politically engaged them. Barnet's increase in turnout and surge to the Conservatives in 2018 contradicted UK trends for regional and demographic voting in remain supporting areas. In the 2017 General Election, all three Barnet constituencies saw increased turnout and increased Labour votes. Since turnout was up in almost every ward in 2018, surely some of these new engaged voters from 2016 or 2017 were voting again in 2018, possibly making the Conservative gains even more remarkable.

The Conservatives success can not be described only as attracting votes from Labour or UKIP, nor Labour's failure excused by their increase in votes. But combining the voting patterns and turnout changes with demographic analysis and anecdotal evidence from those on the campaign trail creates a picture that suggests anti-semitism may have been the over-riding issue that drove voters to the Conservatives in many key Barnet wards in 2018.


